---
title: "Mtcars Project - Automatic or Manual Transmission"
author: "Joe Okelly"
date: "20 Aug 2024"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

In this study we look at the mtcars dataset, which comprises of 32 different automobile design.
we explore the relationship between miles per gallon. We specially focus automatic and manual transmission, and whether MPG (miles per gallon) is better for Manual Transmission cars.

We consider the following point for our analysis :

- Initial Data processing
- Exploratory Data Analysis
- Model Selection
- Technical Analysis 
- Final Analysis
- Conclusion



## Initial Data Preprocessing

We use factor to change all the continuous variable to descrete, ie the 'am' variable which denotes whether a car is automatic or manual. 


```{r}
data("mtcars")
data <- mtcars
data$am <- as.factor(data$am)
levels(data$am) <- c("A", "M") 

data$cyl <- as.factor(data$cyl)
data$gear <- as.factor(data$gear)
data$vs <- as.factor(data$vs)
levels(data$vs) <- c("V", "S")

```

## Exploratory Data Analysis

First we take a look at the data set, we display the first 5 rows of data

```{r}
str(data)
head(data, n = 5)

```

Will will now create a boxplot to visaualize the relationship between mpg and am.

```{r}
library(ggplot2)
g <- ggplot(data, aes(am, mpg))
g <- g + geom_boxplot(aes(fill = am))
print(g)

```
The plot show that cars with Manual transmission have a higher mpg as compared to the cars with Automatic transmission. we might be overlooking some other parameters which has a high correlation with the mpg. 
Lets do more analysis with all these variables whose correlation with mpg is higher.



```{r}
correlation <- cor(mtcars$mpg, mtcars)
correlation <- correlation[,order(-abs(correlation[1, ]))]
correlation

variables <- names(correlation)[1: which(names(correlation) == "am")]
variables
    
```

## Model Selection

Now that we know mpg variable has stronger correlations with other variables too apart from just am, we can't base our model solely on this one variable as it will not be the most accurate one. Let's start this process by fitting mpg with just am.

```{r}
first <- lm(mpg ~ am, data)
summary(first)

```

In this case p-value is quite low but the R-squared value is the real problem. Hence, let's now go to the other extreme end and fit all variables with mpg.

```{r}
last <- lm(mpg ~ ., data)
summary(last)

```

Here R-squared values have definitely improved but the p-value becomes the problem now which is caused most probably due to overfitting. So, lets use 'step' method to iterate over the variables and obtain the best model.

```{r}
best <- step(last, direction = "both", trace = FALSE)
summary(best)

```

Here the R-squared value is pretty good and also p-values are quite significant. Hence undoubtedly this is the best fit for us.

## Model Examination

The best model we obtained i.e., 'best' depicts the dependance of mpg over wt and qsec other than am. Let's plot and study some residual plots to understand more about the 'best' fit. 

```{r}
layout(matrix(c(1,2,3,4),2,2))
plot(best)

```

## Conclusion

The first question whether automatic or manual is better for mpg can be answered using all the models created as holding all the other parameters constant, manual transmission increases the mpg.

However the second question is a little difficult to answer.
Based on 'best' fit model, we conclude that cars with manual transmission have 2.93 more mpg than that of automatic with p < 0.05 and R-squared 0.85.

Residuals vs Fitted plot however shows something is missing from the model which might be a problem due to a small sample size which is 32 observations. Even though the conclusion that manual has better performance with respect to mpg, whether the model will git all future observations will be doubtful.



