---
title: "Mtcars Project - Automatic or Manual Transmission"
author: "Joe Okelly"
date: "20 Aug 2024"
output:
  html_document:
    keep_md: yes
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Executive Summary

In this study we look at the mtcars dataset, which comprises of 32 different automobile design.
we explore the relationship between miles per gallon. We specially focus automatic and manual transmission, and whether MPG (miles per gallon) is better for Manual Transmission cars.


We consider the following point for our analysis :

- Initial Data processing
- Exploratory data Analysis
- Model Selection
- Technical Analysis 
- Final Analysis
- Conclusion


## Initial Data Processing

We use factor to change all the continuous variable to descrete, ie the 'am' variable which denotes whether a car is automatic or manual. 

````{r}
data("mtcars")
data <- mtcars
data$mpg <- as.factor(data$mpg)
data$am <- as.factor(data$am)
levels(data$am) <- c("A", "M") 

data$cyl <- as.factor(data$cyl)
data$gear <- as.factor(data$gear)
data$vs <- as.factor(data$vs)
levels(data$vs) <- c("V", "S")

````

## Exploratiory Data Analysis

First we take a look at the data set, we display the first 5 rows of data

````{r}
str(data)
head(data, n=5)

````

Will will now create a boxplot to visaualize the relationship between mpg and am.

